---
title: "Lab 1: Getting Started with R and First Ecological Simulation"
subtitle: "Creating our first ecological dataset"
author: "Stoyan Stoyanov"
format:
  html:
    toc: true
    number-sections: true
---

# Learning objectives

After completing this lab, you should be able to:

- understand the basic RStudio interface;
- create and manipulate simple R objects;
- use R commands to perform calculations;
- create simple visualizations;
- simulate a small ecological dataset;
- understand the difference between a model and observations.

# Part 1: Getting started with R

## What is R?

R is a programming language and environment for statistical computing.

In this course, we will use R to:

- explore ecological data;
- create simulations;
- fit statistical models;
- visualize ecological patterns.

R is not only a calculator.

It allows us to describe processes and investigate how data are generated.

# RStudio interface

RStudio provides an environment for working with R.

The main components are:

- Console: where commands are executed;
- Script editor: where code is saved;
- Environment: where objects are stored;
- Plots: where visualizations are displayed.

During this course, we will usually write code in scripts rather than directly
in the Console.

# First calculations

R can be used as a calculator.

```{r}
2 + 2
```

Try:

```{r}
10 * 5
```

and:

```{r}
100 / 4
```

# Creating objects

In R, we store information in objects.

Example:

```{r}
age <- 5

age
```

The symbol `<-` assigns a value to an object.

We can create a sequence of values:

```{r}
ages <- 1:10

ages
```

# Creating vectors

A vector is a collection of values.

Example:

```{r}
temperature <- c(12, 15, 18, 20, 22)

temperature
```

We can calculate:

```{r}
mean(temperature)
```

The function `mean()` calculates the average value.

# Basic visualization

Visualization is one of the main tools in ecology.

Example:

```{r}
plot(ages)
```

We will use graphs throughout the course to understand ecological patterns.

# Part 2: Our first ecological simulation

## The ecological question

Imagine that we study a fish population.

Our question is:

> How does fish age influence fish mass?

Our hypothesis:

> Older fish tend to be heavier.

The conceptual model is:

```
Age

 ↓

Growth process

 ↓

Fish mass
```

# Creating a simple growth model

First, we create ages:

```{r}
age <- 1:20
```

Now we create an expected relationship between age and mass:

```{r}
expected_mass <- 20 + 15 * age

expected_mass
```

This represents an ideal population where all fish follow exactly the same
growth pattern.

However, real populations are not like this.

# Adding biological variation

Individuals differ.

We introduce random variation:

```{r}
set.seed(123)

observed_mass <- expected_mass +
  rnorm(length(age), mean = 0, sd = 20)

observed_mass
```

Now we have simulated observations.

The data contain:

- a growth process;
- individual variation;
- random differences.

# Creating our first dataset

We combine the variables:

```{r}
fish_data <- data.frame(
  age = age,
  mass = observed_mass
)

fish_data
```

This is our first ecological dataset.

# Visualizing the population

```{r}
plot(
  fish_data$age,
  fish_data$mass,
  xlab = "Age",
  ylab = "Mass",
  pch = 19
)
```

Observe the graph.

Questions:

- Is there a relationship between age and mass?
- Why are points not exactly on a straight line?
- What represents the variation?

# From simulation to statistical modelling

We know how the data were generated.

We created:

```
Age

 ↓

Growth model

 ↓

Observed mass
```

But imagine that we did not know the growth model.

Then our task would be:

> Can we discover the process that generated these observations?

This is the main question of statistical modelling.

# Exercises

## Exercise 1

Change the amount of variation:

```{r}
sd = 5
```

Compare the result with:

```{r}
sd = 40
```

How does variation influence the dataset?

---

## Exercise 2

Increase the number of fish:

```{r}
age <- 1:100
```

How does a larger sample change your interpretation?

---

## Exercise 3

Discuss:

- Why do ecological datasets contain variation?
- Why is simulation useful?
- What information is lost when we only analyse observations?

# Summary

In this first lab, we learned that:

- R can be used to create and explore data;
- ecological observations are generated by processes;
- variation is a fundamental property of ecological systems;
- simulation helps us understand statistical models.